Fast approximate Bayesian inference for the Naomi HIV surveillance model

### 2023-02-20 src/docs_paper/appendix.Rmd
Alex's feedback round two

Let $\nu$ the most recent national household survey with HIV testing which has taken place in the country of interest at time $t = T_1$
Let $x$ refer to a district located within the Spectrum [@stover2010spectrum] region $R_x$.
Let $s \in \{\text{F}, \text{M}\}$ be sex, and $a \in \{\text{0-5}, \text{5-10}, \ldots, \text{75-80}, 80+\}$ be five-year age bands.
As short-hand, we write $a = l$ to refer to the age band with lower bound $l$, e.g. $a = 20$ for $a = \text{20-25}$.
We index the following quantities by district, sex and age-band: population size $N_{x, s, a}$, HIV prevalence $\rho_{x, s, a}$, ART coverage $\alpha_{x, s, a}$, annual HIV incidence rate $\lambda_{x, s, a}$, and proportion of HIV positive persons recently infected $\kappa_{x, s, a}$.
Sometimes data are observed at an aggregate level, rather than the more granular modelled level.
In this instance, we use $\{\cdot\}$ to generically refer to a aggregate set over which an observation is made, e.g. $\{a\} = \{\text{15-19}, \ldots, \text{45-49}\}$ for adults.

>>>>

Let $\nu$ the most recent national household survey with HIV testing which has taken place in the country of interest at time $t = T_1$
Let $x \in \{1, \ldots, n\}$ refer to a district located within the Spectrum [@stover2010spectrum] region $R_x$.
Let $s \in \{\text{F}, \text{M}\}$ be sex, and $a \in \{\text{0-5}, \text{5-10}, \ldots, \text{75-80}, \text{80+}\}$ be five-year age bands.
As short-hand, we write $a = l$ to refer to the age band with lower bound $l$, e.g. $a = 20$ for $a = \text{20-25}$.
We index the following quantities by district, sex and age-band: population size $N_{x, s, a}$, HIV prevalence $\rho_{x, s, a}$, ART coverage $\alpha_{x, s, a}$, annual HIV incidence rate $\lambda_{x, s, a}$, and proportion of HIV positive persons recently infected $\kappa_{x, s, a}$.
Sometimes data are observed at an aggregate level, rather than the more granular modelled level.
In this instance, we use $\{\cdot\}$ to generically refer to a aggregate set over which an observation is made, e.g. $\{a\} = \{\text{15-19}, \ldots, \text{45-49}\}$ for adults.

### 2023-02-20 src/docs_paper/paper.Rmd
Alex's feedback round two

---
title: Integrated nested Laplace approximations for extended latent Gaussian models with application to the Naomi HIV model
author:
  - name: Adam 
    surname: Howes
    email: ath19@ic.ac.uk
    label: e1
    addressLabel: A,D
    sepNext: ","
  - name: Alex
    surname: Stringer
    email: alex.stringer@uwaterloo.ca
    label: e2
    addressLabel: B
  - name: Seth R.
    surname: Flaxman
    email: seth.flaxman@cs.ox.ac.uk
    label: e3
    addressLabel: C
    sepNext: ","
  - name: Jeffrey W.
    surname: Eaton
    email: jeffrey.eaton@imperial.ac.uk
    label: e4
    addressLabel: D
affiliation:
  - label: A
    name: Department of Mathematics, Imperial College London
    authorsLabels: e1
  - label: B
    name: Department of Statistics and Actuarial Science, University of Waterloo
    authorsLabels: e2
  - label: C
    name: Department of Computer Science, University of Oxford
    authorsLabels: e3
  - label: D
    name: MRC Centre for Global Infectious Disease Analysis, School of Public Health, Imperial College London
    authorsLabels: e4
abstract: |
  | Naomi is a spatial evidence synthesis model used to produce district-level HIV epidemic indicators in sub-Saharan Africa. Multiple outcomes of policy interest, including HIV prevalence, HIV incidence and antiretroviral therapy treatment coverage, are jointly modelled using both household survey data and routinely reported health system data. Inference for Naomi is currently conducted using an empirical Bayes Gaussian approximation via the \texttt{TMB} \textsc{R} package. We propose a new inference method which combines the simplified integrated nested Laplace approximation approach of @wood2020simplified with adaptive Gauss-Hermite quadrature to enable fast and accurate inference for Naomi and other extended latent Gaussian models. Using data from Malawi, our method provides more accurate inferences than \texttt{TMB}, and is comparable to Hamiltonian Monte Carlo with the No-U-Turn sampler, but faster to run. By extending the \texttt{aghq} \textsc{R} package we facilitate easy, flexible use of our method when provided a \texttt{TMB} \textsf{C++} template for the model's log-posterior. In doing so, we enable inference via integrated nested Laplace approximations for a larger class of models than was previously possible.
keyword-subclass: | 
 \begin{keyword}[class=MSC2020] % It must be define for aap, aop, aos journals. For aoas, sts is not used
 \kwd[Primary ]{00X00}
 \kwd{00X00}
 \kwd[; secondary ]{00X00}
 \end{keyword}
keywords: 
  - spatial statistics
  - small-area estimation
  - INLA
  - AGHQ
  - HIV epidemiology

@@

@eaton2021naomi specify a joint model linking three small-area estimation models.
The full model is defined over three time points: $T_1$ the time of the most recent household survey with HIV testing; $T_2$, the current time period; and $T_3$, a short term projection period.
We consider a simplified version defined only at $T_1$ omitting nowcasting of $T_2$ and temporal projection to $T_3$.
Below we provide an overview of the simplified model, highlighting the aspects which make it a challenge for existing inferential approaches.
A more complete mathematical description of the simplified model, as well as a \textsf{C++} template for the log-posterior, are provided in the appendix.

@@

Consider a country in sub-Saharan Africa where a household survey with complex survey design has taken place at time $T_1$.
Let $x$ index district, $a$ five-year age band, and $s$ sex.
For ease of notation, let $i$ index the finest district-age-sex division included in the model.
As data we observe may be aggregated over indices $i$, let $\mathcal{I}$ be a set of $i$ for which observations are reported.
Let $N_i \in \mathbb{N}$ be the population size, $\rho_i \in [0, 1]$ be HIV prevalence, $\alpha_i \in [0, 1]$ be antiretroviral therapy (ART) coverage, and $\lambda_i > 0$ be the annual HIV incidence rate.
Independent mixed effects models for HIV prevalence and ART coverage in the general population are specified on the logit scale such that
\begin{align*}
\text{logit}(\rho_i) &= \eta^\rho_i, \\
\text{logit}(\alpha_i) &= \eta^\alpha_i,
\end{align*}
for certain choice of linear predictors $\eta^\rho_i$ and $\eta^\alpha_i$.
For the HIV incidence rate we use a mixed effects model on the log scale $\log(\lambda_i) = \eta^\lambda_i(\{\rho_i, \alpha_i\}_{i \in \mathcal{I}})$, where the linear predictor depends on $\{\rho_i, \alpha_i\}_{i \in \mathcal{I}}$ for some $\mathcal{I}$.
Let $\kappa_i$ be the proportion recently infected among HIV positive persons.
<!-- The kappa is introduced, but you have to be paying close attention to see that it's so that for theta in rho, alpha, kappa -->
For each set of observed strata indices $\mathcal{I}$, we calculate the weighted observations $\hat \theta_\mathcal{I}$ for $\theta \in \{\rho, \alpha, \kappa\}$ with respective Kish effective sample sizes
$$
m^{\hat \theta}_\mathcal{I} = \frac{\left(\sum_j w_j\right)^2}{\sum_j w_j^2},
$$
where $j$ indexes individuals across all strata $i \in \mathcal{I}$.
The corresponding design weights are $w_j = 1 / \pi_j$, where $\pi_j$ is the probability of inclusion in the survey.
The observed number of indicator cases is then $y^{\hat \theta}_{\mathcal{I}} = m^{\hat \theta}_{\mathcal{I}} \cdot \hat \theta_{\mathcal{I}}$.
For $\theta \in \{\rho, \alpha, \kappa\}$ we model these aggregate observations using a binomial working likelihood $y^{\hat \theta}_{\mathcal{I}} \sim \text{xBin}(m^{\hat \theta}_{\mathcal{I}}, \theta_{\mathcal{I}})$, where $\theta_{\mathcal{I}}$ are the following appropriately weighted aggregates
\begin{align*}
\rho_{\mathcal{I}} = \frac{\sum_{i \in \mathcal{I}} N_i \rho_i}{\sum_{i \in \mathcal{I}} N_i}, \\
\alpha_{\mathcal{I}} = \frac{\sum_{i \in \mathcal{I}} N_i \rho_i \alpha_i}{\sum_{i \in \mathcal{I}} N_i \rho_i}, \\
\kappa_{\mathcal{I}} = \frac{\sum_{i \in \mathcal{I}} N_i \rho_i \kappa_i}{\sum_{i \in \mathcal{I}} N_i \rho_i}.
\end{align*}
To link the proportion recently infected among HIV positive persons $\kappa_i$ to HIV incidence $\lambda_i$ we use
$$
\kappa_i = 1- \exp \left( - \lambda_i \cdot \frac{1 - \rho_i}{\rho_i} \cdot (\Omega_T - \beta_T) - \beta_T \right),
$$
where the mean duration of recent infection $\Omega_T$ and false recent ratio $\beta_T$ are strongly informed by priors for the particular survey.

@@

Extended latent Gaussian models (ELGMs) [@stringer2022fast] relax the restriction that there is a one-to-one mapping between the mean response $\bmu$ and structured additive predictor $\bmeta$.
Instead, the structured additive predictor is redefined as $\bmeta = (\eta)_{i \in [N_n]}$, where $N_n \in \mathbb{N}$ depends on, but may not be equal to, $n$.
Each mean response $\mu_i$ now depends on some subset $\mathcal{J}_i \subseteq [N_n]$ of indices of $\bmeta$, with $\cup_{i = 1}^n \mathcal{J}_i = [N_n]$ and $1 \leq |\mathcal{J}_i| \leq N_n$.
The inverse link function $g(\cdot)$ is redefined for each observation to be a possibly many-to-one mapping $g_i: \mathbb{R}^{|\mathcal{J}_i|} \to \mathbb{R}$, such that $\mu_i = g_i(\bmeta_{\mathcal{J}_i})$.
Importantly, this allows for the presence of more non-linearity.
ELGMs are then of the form
\begin{align*}
y_i &\sim p(y_i \, | \, \bmeta_{\mathcal{J}_i}, \btheta_1), \quad i \in [n] \\
\mu_i &= \mathbb{E}(y_i \, | \, \bmeta_{\mathcal{J}_i}) = g_i(\bmeta_{\mathcal{J}_i}), \\
\eta_j &= \beta_0 + \sum_{l = 1}^{p} \beta_j z_{ji} + \sum_{k = 1}^{r} f_k(u_{ki}), \quad j \in [N_n].
\end{align*}

@@

The joint posterior of $(\x, \btheta)$ for an ELGM is given by
\begin{equation*}
  p(\x, \btheta \, | \, \y)
  \propto p(\btheta) |\mathbf{Q}(\btheta)|^{n/2} \exp \left( - \frac{1}{2} \x^\top \mathbf{Q}(\btheta) \x + \sum_{i = 1}^n \log p(y_i \, | \, x_i, \btheta) \right).
\end{equation*}
We consider approximations to the posterior marginals of each latent random variable $x_i$ and hyperparameter $\theta_j$ given by
\begin{align}
  \tilde p(x_i \, | \, \y) \approx p(x_i \, | \, \y) &= \int p(x_i, \btheta \, | \, \y) \text{d} \btheta = \int p(x_i \, | \, \btheta, \y) p(\btheta \, | \, \y) \text{d}\btheta, \quad i \in [N], \label{eq:inla1} \\
  \tilde p(\theta_j \, | \, \y) \approx p(\theta_j \, | \, \y) &= \int p(\btheta \, | \, \y) \text{d}\btheta_{-j} \quad j \in [m]. \label{eq:inla2}
\end{align}

@@

<!-- To-do: define what exactly $p(\y, \x, \btheta)$ is -->
Given a \textsf{C++} user template `model.cpp` for the negative unnormalised log posterior $- \log p(\y, \x, \btheta)$, we obtain the posterior marginal approximations $\{ \tilde p(x_i \, | \, \y) \}_{i = 1}^n$ and $\tilde p(\theta_j \, | \, \y)_{j = 1}^m$ via the following algorithm, comprised of nested applications of Laplace approximation and adaptive Gauss-Hermite quadrature.

@@

1. Calculate the mode, Hessian at the mode, and lower Cholesky
\begin{align*}
\hat{\btheta} &= \argmax_{\btheta} {\tilde p_\text{LA}(\btheta, \y)}, \\
\mathbf{H} &= \frac{\partial^2}{\partial \btheta \partial \btheta^\top} - \log \tilde p_\text{LA}(\btheta, \y) \rvert_{\btheta = \hat \btheta} \\
\mathbf{H}^{-1} &= \mathbf{L} \mathbf{L}^\top,
\end{align*}
of the Laplace approximation
$$
\tilde p_\text{LA}(\btheta, \y) = \frac{p(\y, \x, \btheta)}{\tilde p_\text{G}(\x \, | \, \btheta, \y)} \Big\rvert_{\x = \hat \x(\btheta)}
$$
where $\tilde p_\text{G}(\x \, | \, \btheta, \y) = \mathcal{N}(\x \, | \, \hat \x(\btheta), \mathbf{H}(\btheta)^{-1})$ is a Gaussian approximation to $p(\x \, | \, \btheta, \y)$ with mode and precision matrix given by
\begin{align*}
\hat \x(\btheta) &= \argmin_\x - \log p(\y, \x, \btheta), \\
\mathbf{H}(\btheta) &= \frac{\partial^2}{\partial x \partial x^\top} - \log p(\y, \x, \btheta) \rvert_{\x = \hat \x(\btheta)}.
\end{align*}
2. Generate a set of nodes $\z \in \mathcal{Q}(m, k)$ and weights $\omega: \z \in \mathcal{Q}(m, k) \to \mathbb{R}$ from a Gauss-Hermite quadrature rule with $k$ nodes per dimension, which are then adapted based on the mode and lower Choleksy via $\btheta(\z) = \hat{\btheta} + \mathbf{L} \z$.
If possible $k \geq 3$ is prefered, though the number of grid points scales exponentially with choice of $k$.
3. Use this quadrature rule to calculate the normalising constant $\tilde p_{\text{AQ}}(\y)$ as follows
\begin{equation}
\tilde p_{\text{AQ}}(\y) = \sum_{\z \in \mathcal{Q}(m, k)} \tilde p_\text{LA}(\btheta(\z), \y) \omega(\z). \label{eq:evidence1}
\end{equation}
4. For $i \in [n]$ generate $l$ nodes $x_i(\z)$, via a Gauss-Hermite quadrature rule $\z \in \mathcal{Q}(1, l)$ adapted based on the mode $\hat \x(\btheta)_i$ and standard deviation $\sqrt{\text{diag}[\mathbf{H}(\btheta)^{-1}]_i}$ of the Gaussian marginal. A value of $l \geq 4$ is recommended to enable B-spline interpolation.
5. For $x_i \in \{ x_i(\z) \}_{\z \in \mathcal{Q}(1, l)}$  and $\btheta \in \{ \btheta(\z) \}_{\z \in \mathcal{Q}(m, k)}$ calculate the modes and Hessians
\begin{align*}
\hat \x_{-i}(x_i, \btheta) &= \argmin_{\x_{-i}} - \log p(\y, x_i, \x_{-i}, \btheta), \\
\mathbf{H}_{-i, -i}(x_i, \btheta) &= \frac{\partial^2}{\partial \x_{-i} \partial \x_{-i}^\top} - \log p(\y, x_i, \x_{-i}, \btheta) \rvert_{\x_{-i} = \hat \x_{-i}(x_i, \btheta)},
\end{align*}
where optimisation to obtain $\hat \x_{-i}(x_i, \btheta)$ is initialised at $\hat \x(\btheta)_{-i}$.
6. For $x_i \in \{ x_i(\z) \}_{\z \in \mathcal{Q}(1, l)}$ calculate
\begin{equation}
\tilde p_\text{AQ}(x_i \, | \, \y) = \frac{\tilde p_\text{LA}(x_i, \y)}{\tilde p_{\text{AQ}}(\y)}. \label{eq:laplace-marginal}
\end{equation}
where
$$
\tilde p_\text{LA}(x_i, \y) = \sum_{\z \in \mathcal{Q}(m, k)} \tilde p_\text{LA}(x_i, \btheta(\z), \y) \omega(\z).
$$
and
$$
\tilde p_\text{LA}(x_i, \btheta, \y) = \frac{p(x_i, \x_{-i}, \btheta, \y)}{\tilde p_\text{G}(\x_{-i} \, | \, x_i, \btheta, \y)} \Big\rvert_{\x_{-i} = \hat \x_{-i}(x_i, \btheta)}.
$$
Although Equation \ref{eq:laplace-marginal} can be calculated using the estimate of the evidence in Equation \ref{eq:evidence1} it is more numerically accurate to use the estimate
$$
\tilde p_{\text{AQ}}(\y) = \sum_{\z \in \mathcal{Q}(1, l)} \tilde p_\text{LA}(x_i(\z), \y) \omega(\z)
$$

>>>>

---
title: Fast approximate Bayesian inference for the Naomi HIV surveillance model
author:
  - name: Adam 
    surname: Howes
    email: ath19@ic.ac.uk
    label: e1
    addressLabel: A,D
    sepNext: ","
  - name: Alex
    surname: Stringer
    email: alex.stringer@uwaterloo.ca
    label: e2
    addressLabel: B
  - name: Seth R.
    surname: Flaxman
    email: seth.flaxman@cs.ox.ac.uk
    label: e3
    addressLabel: C
    sepNext: ","
  - name: Jeffrey W.
    surname: Eaton
    email: jeffrey.eaton@imperial.ac.uk
    label: e4
    addressLabel: D
affiliation:
  - label: A
    name: Department of Mathematics, Imperial College London
    authorsLabels: e1
  - label: B
    name: Department of Statistics and Actuarial Science, University of Waterloo
    authorsLabels: e2
  - label: C
    name: Department of Computer Science, University of Oxford
    authorsLabels: e3
  - label: D
    name: MRC Centre for Global Infectious Disease Analysis, School of Public Health, Imperial College London
    authorsLabels: e4
abstract: |
  | Naomi is a spatial evidence synthesis model used to produce district-level HIV epidemic indicators in sub-Saharan Africa. Multiple outcomes of policy interest, including HIV prevalence, HIV incidence and antiretroviral therapy treatment coverage, are jointly modelled using both household survey data and routinely reported health system data. Inference for Naomi is currently conducted using an empirical Bayes Gaussian approximation via the \texttt{TMB} \textsc{R} package. We propose a new inference method which combines the simplified integrated nested Laplace approximation approach of @wood2020simplified with adaptive Gauss-Hermite quadrature to enable fast and accurate inference for Naomi and other extended latent Gaussian models. Using data from Malawi, our method provides more accurate inferences than \texttt{TMB}, and is comparable to Hamiltonian Monte Carlo with the No-U-Turn sampler, but faster to run. By extending the \texttt{aghq} \textsc{R} package we facilitate easy, flexible use of our method when provided a \texttt{TMB} \textsf{C++} template for the model's log-posterior. In doing so, we enable inference via integrated nested Laplace approximations for a larger class of models than was previously possible.
keyword-subclass: | 
 \begin{keyword}[class=MSC2020] % It must be define for aap, aop, aos journals. For aoas, sts is not used
 \kwd[Primary ]{00X00}
 \kwd{00X00}
 \kwd[; secondary ]{00X00}
 \end{keyword}
keywords: 
  - spatial statistics
  - small-area estimation
  - INLA
  - AGHQ
  - HIV epidemiology

@@

@eaton2021naomi specify a joint model linking three small-area estimation models.
The model is defined over three time points: $T_1$ the time of the most recent household survey with HIV testing; $T_2$, the current time period; and $T_3$, a short term projection period.
We consider a simplified version defined only at $T_1$ omitting nowcasting of $T_2$ and temporal projection to $T_3$.
Below we provide an overview of the simplified model, highlighting the aspects which make it a challenge for existing inferential approaches.
A more complete mathematical description of the simplified model, as well as a \textsf{C++} template for the log-posterior, are provided in the appendix.

@@

Consider a country in sub-Saharan Africa where a household survey with complex survey design has taken place at time $T_1$.
Let $x \in \mathcal{X}$ index district, $a \in \mathcal{A}$ five-year age band, and  $s \in \mathcal{S}$ sex.
For ease of notation, let $i$ index the finest district-age-sex division included in the model.
The data we observe may be aggregated over indices $i$.
Let $\mathcal{I} \subseteq \mathcal{X} \times \mathcal{A} \times \mathcal{S}$ be a set of  indicies $i$ for which an observation is reported.

Let $N_i \in \mathbb{N}$ be the known, fixed population size.
We infer the following unknown HIV indicators: HIV prevalence $\rho_i \in [0, 1]$, the proporiton of indviduals who are HIV positive; antiretroviral therapy (ART) coverage $\alpha_i \in [0, 1]$, the proportion of people living with HIV who receive ART treatment; and annual HIV incidence rate $\lambda_i > 0$, the yearly rate of new HIV infections occurring.
Independent logistic regression models for HIV prevalence and ART coverage in the general population are specified on such that
\begin{align*}
\text{logit}(\rho_i) &= \eta^\rho_i, \\
\text{logit}(\alpha_i) &= \eta^\alpha_i,
\end{align*}
for certain choice of linear predictors $\eta^\rho_i$ and $\eta^\alpha_i$.
For the HIV incidence rate we model on the log scale $\log(\lambda_i) = \eta^\lambda_i(\{\rho_i, \alpha_i\}_{i \in \mathcal{I}})$, where the linear predictor depends on $\{\rho_i, \alpha_i\}_{i \in \mathcal{I}}$ for some $\mathcal{I}$.
Let $\kappa_i$ be the proportion recently infected among HIV positive persons, which we link to HIV incidence via
$$
\kappa_i = 1- \exp \left( - \lambda_i \cdot \frac{1 - \rho_i}{\rho_i} \cdot (\Omega_T - \beta_T) - \beta_T \right),
$$
where the mean duration of recent infection $\Omega_T$ and false recent ratio $\beta_T$ are strongly informed by priors for the particular survey.

For $\theta \in \{\rho, \alpha, \kappa\}$ we calculate the weighted observations
$$
\hat \theta_\mathcal{I} = \frac{\sum_j w_j \cdot\theta_j}{\sum_j w_j},
$$
where $j$ indexes individuals across all strata $i \in \mathcal{I}$.
The design weights are
$$
w_j = \frac{1}{\pi_j} \times \frac{1}{\omega_j},
$$
where $\pi_j$ is the probability of inclusion and $\omega_j$ is a non-response factor for the particular survey.
We calculate the observed number of indicator cases as $y^{\hat \theta}_{\mathcal{I}} = m^{\hat \theta}_{\mathcal{I}} \cdot \hat \theta_{\mathcal{I}}$ where
$$
m^{\hat \theta}_\mathcal{I} = \frac{\left(\sum_j w_j\right)^2}{\sum_j w_j^2},
$$
is the Kish effective sample size [@kish1965survey].

For $\theta \in \{\rho, \alpha, \kappa\}$ we model these aggregate observations using a binomial working likelihood $y^{\hat \theta}_{\mathcal{I}} \sim \text{xBin}(m^{\hat \theta}_{\mathcal{I}}, \theta_{\mathcal{I}})$, where $\theta_{\mathcal{I}}$ are the following appropriately weighted aggregates
\begin{align*}
\rho_{\mathcal{I}} = \frac{\sum_{i \in \mathcal{I}} N_i \rho_i}{\sum_{i \in \mathcal{I}} N_i}, \\
\alpha_{\mathcal{I}} = \frac{\sum_{i \in \mathcal{I}} N_i \rho_i \alpha_i}{\sum_{i \in \mathcal{I}} N_i \rho_i}, \\
\kappa_{\mathcal{I}} = \frac{\sum_{i \in \mathcal{I}} N_i \rho_i \kappa_i}{\sum_{i \in \mathcal{I}} N_i \rho_i}.
\end{align*}

@@

Extended latent Gaussian models (ELGMs) [@stringer2022fast] relax the restriction that there is a one-to-one mapping between the mean response $\bmu$ and structured additive predictor $\bmeta$.
Instead, the structured additive predictor is redefined as $\bmeta = (\eta)_{i \in [N_n]}$, where $N_n \in \mathbb{N}$ is a function of $n$, and it is possible that $N_n \neq n$.
Each mean response $\mu_i$ now depends on some subset $\mathcal{J}_i \subseteq [N_n]$ of indices of $\bmeta$, with $\cup_{i = 1}^n \mathcal{J}_i = [N_n]$ and $1 \leq |\mathcal{J}_i| \leq N_n$.
The inverse link function $g(\cdot)$ is redefined for each observation to be a possibly many-to-one mapping $g_i: \mathbb{R}^{|\mathcal{J}_i|} \to \mathbb{R}$, such that $\mu_i = g_i(\bmeta_{\mathcal{J}_i})$.
Importantly, this allows for the presence of more non-linearity.
ELGMs are then of the form
\begin{align*}
y_i &\sim p(y_i \, | \, \bmeta_{\mathcal{J}_i}, \btheta_1), \quad i \in [n] \\
\mu_i &= \mathbb{E}(y_i \, | \, \bmeta_{\mathcal{J}_i}) = g_i(\bmeta_{\mathcal{J}_i}), \\
\eta_j &= \beta_0 + \sum_{l = 1}^{p} \beta_j z_{ji} + \sum_{k = 1}^{r} f_k(u_{ki}), \quad j \in [N_n].
\end{align*}

@@

The joint posterior of $(\x, \btheta)$ for an ELGM is given by
\begin{equation*}
  p(\x, \btheta \, | \, \y)
  \propto p(\btheta) |\mathbf{Q}(\btheta)|^{n/2} \exp \left( - \frac{1}{2} \x^\top \mathbf{Q}(\btheta) \x + \sum_{i = 1}^n \log p(y_i \, | \, \x_{\mathcal{J}_i}, \btheta) \right).
\end{equation*}
We consider approximations to the posterior marginals of each latent random variable $x_i$ and hyperparameter $\theta_j$ given by
\begin{align}
  \tilde p(x_i \, | \, \y) \approx p(x_i \, | \, \y) &= \int p(x_i, \btheta \, | \, \y) \text{d} \btheta = \int p(x_i \, | \, \btheta, \y) p(\btheta \, | \, \y) \text{d}\btheta, \quad i \in [N], \label{eq:inla1} \\
  \tilde p(\theta_j \, | \, \y) \approx p(\theta_j \, | \, \y) &= \int p(\btheta \, | \, \y) \text{d}\btheta_{-j} \quad j \in [m]. \label{eq:inla2}
\end{align}

@@

<!-- To-do: define what exactly $p(\y, \x, \btheta)$ is -->
Given the negative unnormalised log posterior $- \log p(\y, \x, \btheta)$, we obtain the posterior marginal approximations $\{ \tilde p(x_i \, | \, \y) \}_{i = 1}^n$ and $\tilde p(\theta_j \, | \, \y)_{j = 1}^m$ via the following algorithm, comprised of nested applications of Laplace approximation and adaptive Gauss-Hermite quadrature.

@@

1. Calculate the mode, Hessian at the mode, and lower Cholesky
\begin{align*}
\hat{\btheta} &= \argmax_{\btheta} {\tilde p_\text{LA}(\btheta, \y)}, \\
\mathbf{H} &= \frac{\partial^2}{\partial \btheta \partial \btheta^\top} - \log \tilde p_\text{LA}(\btheta, \y) \rvert_{\btheta = \hat \btheta} \\
\mathbf{H}^{-1} &= \mathbf{L} \mathbf{L}^\top,
\end{align*}
of the Laplace approximation
$$
\tilde p_\text{LA}(\btheta, \y) = \frac{p(\y, \x, \btheta)}{\tilde p_\text{G}(\x \, | \, \btheta, \y)} \Big\rvert_{\x = \hat \x(\btheta)}
$$
where $\tilde p_\text{G}(\x \, | \, \btheta, \y) = \mathcal{N}(\x \, | \, \hat \x(\btheta), \mathbf{H}(\btheta)^{-1})$ is a Gaussian approximation to $p(\x \, | \, \btheta, \y)$ with mode and precision matrix given by
\begin{align*}
\hat \x(\btheta) &= \argmin_\x - \log p(\y, \x, \btheta), \\
\mathbf{H}(\btheta) &= \frac{\partial^2}{\partial x \partial x^\top} - \log p(\y, \x, \btheta) \rvert_{\x = \hat \x(\btheta)}.
\end{align*}
2. Generate a set of nodes $\z \in \mathcal{Q}(m, k)$ and weights $\omega: \z \in \mathcal{Q}(m, k) \to \mathbb{R}$ from a Gauss-Hermite quadrature rule with $k$ nodes per dimension, which are then adapted based on the mode and lower Choleksy via $\btheta(\z) = \hat{\btheta} + \mathbf{L} \z$.
If possible $k \geq 3$ is preferred, though the number of grid points scales exponentially with choice of $k$.
Then use this quadrature rule to calculate the normalising constant $\tilde p_{\text{AQ}}(\y)$ as follows
\begin{equation}
\tilde p_{\text{AQ}}(\y) = \sum_{\z \in \mathcal{Q}(m, k)} \tilde p_\text{LA}(\btheta(\z), \y) \omega(\z). \label{eq:evidence1}
\end{equation}
4. For $i \in [n]$ generate $l$ nodes $x_i(\z)$ via a Gauss-Hermite quadrature rule $\z \in \mathcal{Q}(1, l)$ adapted based on the mode $\hat \x(\btheta)_i$ and standard deviation $\sqrt{\text{diag}[\mathbf{H}(\btheta)^{-1}]_i}$ of the Gaussian marginal. A value of $l \geq 4$ is recommended to enable B-spline interpolation.
Then, for $x_i \in \{ x_i(\z) \}_{\z \in \mathcal{Q}(1, l)}$  and $\btheta \in \{ \btheta(\z) \}_{\z \in \mathcal{Q}(m, k)}$ calculate the modes and Hessians
\begin{align*}
\hat \x_{-i}(x_i, \btheta) &= \argmin_{\x_{-i}} - \log p(\y, x_i, \x_{-i}, \btheta), \\
\mathbf{H}_{-i, -i}(x_i, \btheta) &= \frac{\partial^2}{\partial \x_{-i} \partial \x_{-i}^\top} - \log p(\y, x_i, \x_{-i}, \btheta) \rvert_{\x_{-i} = \hat \x_{-i}(x_i, \btheta)},
\end{align*}
where optimisation to obtain $\hat \x_{-i}(x_i, \btheta)$ is initialised at $\hat \x(\btheta)_{-i}$.
6. For $x_i \in \{ x_i(\z) \}_{\z \in \mathcal{Q}(1, l)}$ calculate
\begin{equation}
\tilde p_\text{AQ}(x_i \, | \, \y) = \frac{\tilde p_\text{LA}(x_i, \y)}{\tilde p_{\text{AQ}}(\y)}. \label{eq:laplace-marginal}
\end{equation}
where
$$
\tilde p_\text{LA}(x_i, \y) = \sum_{\z \in \mathcal{Q}(m, k)} \tilde p_\text{LA}(x_i, \btheta(\z), \y) \omega(\z).
$$
and
$$
\tilde p_\text{LA}(x_i, \btheta, \y) = \frac{p(x_i, \x_{-i}, \btheta, \y)}{\tilde p_\text{G}(\x_{-i} \, | \, x_i, \btheta, \y)} \Big\rvert_{\x_{-i} = \hat \x_{-i}(x_i, \btheta)}.
$$
Although Equation \ref{eq:laplace-marginal} can be calculated using the estimate of the evidence in Equation \ref{eq:evidence1} it is more numerically accurate to use the estimate
$$
\tilde p_{\text{AQ}}(\y) = \sum_{\z \in \mathcal{Q}(1, l)} \tilde p_\text{LA}(x_i(\z), \y) \omega(\z)
$$
